Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-08-04 manuscripts/reply/reply-abstract.tex
More editing of abstract and text of reply ms.

Establishing that a set of population splitting events occurred at the same
time can be a potentially persuasive argument that the populations were
affected by the same geographic event.
Recently, \citet{Oaks2012} used an approximate-Bayesian model choice method
(\msb) to estimate the support for such a pattern of divergences across 22
pairs of taxa from the Philippines.
They also studied the behavior of \msb using computer simulations.
\citet{Oaks2012} found the method was very sensitive to prior assumptions and
had low power to detect variation in divergence times.
These results were not surprising in light of a rich statistical literature
showing that the marginal likelihood of a model is sensitive to the
use of vague priors.
Because this sensitivity to prior assumptions affects the crucial insights that
a researcher who employs \msb seeks to gain, \citet{Oaks2012} recommended that
users of the approach should carefully assess the robustness of their
conclusions to different priors.
In response, \citet{Hickerson2013} suggested the lack of robustness in \msb
analyses was due to the method's inefficient rejection algorithm and broad
priors on divergence times leading to inadequate numbers of simulation
replicates.
They proposed a model-averaging approach that uses narrow, empirically
informed uniform priors.
Here we demonstrate that the approach of \citet{Hickerson2013} will often
erroneously support shared divergence events and is dangerous in the sense that
the empirically-derived uniform priors often exclude from consideration the
true values of the models' parameters.
On a more fundamental level, we question the value of adopting an empirical
Bayesian stance for this model-choice problem.
The robust Bayesian approach of conducting analyses under a variety of priors
can reveal prior sensitivity and communicate which assumptions underlie model
inference.
% Furthermore, simulations provide insight into the temporal resolution of the
% method, which helps guide interpretation of results.
As predicted by fundamental principles of Bayesian model choice, more flexible
distributions can accommodate prior uncertainty about the models' parameters
while increasing the method's robustness and power to detect temporal variation
in divergences \citep{Oaks2014dpp}.

>>>>

Establishing that a set of population splitting events occurred at the same
time can be a potentially persuasive argument that the populations were
affected by the same geographic event.
Recently, \citet{Oaks2012} assessed the ability of an approximate-Bayesian
model choice method (\msb) to estimate such patterns of divergences across
taxa, to which \citet{Hickerson2013} responded.
Both papers agree that the primary inference enabled by the method is very
sensitive to prior assumptions and often erroneously supports shared
divergences across taxa.
However, the papers differ in the best explanation and solution for this
problem.
\citet{Hickerson2013} suggested the lack of robustness in \msb analyses is
caused by numerical approximation error due to the inefficiency of the method's
rejection algorithm under broad uniform priors on divergence times
(Hypothesis~\ref{hypError}).
To overcome this, they proposed a model-averaging approach that uses narrow,
empirically informed uniform priors.
Here we demonstrate the approach of \citet{Hickerson2013} will often
erroneously support shared divergence events and is dangerous in the sense that
the empirically-derived uniform priors often exclude from consideration the
true values of the divergence-time parameters.
\citet{Oaks2012} suggested the method's behavior was due to the sensitivity of
marginal likelihoods to the uniformly distributed priors on divergence times
(Hypothesis~\ref{hypML}).
To mitigate this sensitivity, they suggested alternative probability
distributions to accommodate prior uncertainty about divergence times.
As predicted by fundamental principles of Bayesian model choice, more flexible
distributions can accommodate prior uncertainty about parameters without
placing excessive prior weight in vast regions of parameter space with low
likelihood;
this increases the method's robustness and power to detect temporal variation
in divergences \citep{Oaks2014dpp}.
We also show that the tendency of \msb analyses to support models of shared
divergences is primarily due to Hypothesis~\ref{hypML};
Hypothesis~\ref{hypError} is not a tenable explanation for the bias.
All of these papers have highlighted the importance of choosing prior
assumptions about parameters in Bayesian model selection, because they can
strongly influence the posterior probabilities of the models we seek to
estimate.
No amount of computation can rescue our inference if we choose priors on
parameters such that the exact posterior supports the wrong model of
evolutionary history.

### 2014-08-04 manuscripts/reply/reply-main.tex
More editing of abstract and text of reply ms.

\begin{linenomath}
Whereas empirical Bayesian approaches can provide powerful methods for
parameter estimation, a theoretical justification for empirical Bayesian
approaches to model choice is questionable.
In Bayesian model choice, the primary goal is not to estimate parameters, but
to estimate the probabilities of candidate models.
In a simple example where two candidate models, $M_1$ and $M_2$, are being
compared, the goal is to estimate the posterior probabilities of these two
models.
Again, we can use Bayes' rule to calculate this as
\begin{equation}
    p(M_1 \given \alignment{i}{}) = \frac{p(\alignment{i}{} \given
    M_1)p(M_1)}{p(\alignment{i}{} \given M_1)p(M_1) + p(\alignment{i}{} \given
    M_2)p(M_2)}.
    \label{eq:bayesmodelchoice}
\end{equation}
By comparing Equations \ref{eq:bayesrule} and \ref{eq:bayesmodelchoice}, we
see fundamental differences between Bayesian parameter estimation and
model choice.
\end{linenomath}

@@

The strong preference for the narrowest prior on divergence times suggests the
approach of \citet{Hickerson2013} is biased toward models with less parameter
space and, as a consequence, will estimate model-averaged posteriors dominated
by models that exclude true values of the parameters.
We explored this possibility in two ways.
First, we re-analyzed the Philippines dataset using the model-averaging
approach of \citet{Hickerson2013}, but set one of the prior models with a
uniform prior on divergence times that is unrealistically narrow and almost
certainly excludes most, if not all, of the true divergence times of the 22
taxon pairs.
If small likelihoods of large models cause the method to prefer models with
less parameter space (Hypothesis~\ref{hypML}), we expect \msb will
preferentially sample from this erroneous model yielding a posterior that is
misleading (i.e., the model-averaged posterior will be dominated by a model
that excludes the truth).
Second, we generated simulated datasets for which the divergence times are
drawn from an exponential distribution and applied the approach of
\citet{Hickerson2013} to each of them to see how often the method excludes the
truth.

@@

\section{The importance of power analyses to guide applications of \texttt{msBayes}}
\citet{Hickerson2013} presented a power analysis of \msb under a narrow uniform
divergence-time prior of 0--1 coalescent units ago.
They found that under these prior conditions \msb can, assuming a
per-site rate of $1.92\e{-8}$ mutations per generation, detect multiple
divergence events among 18 taxa when the true divergences were random over
hundreds of thousands of generations or more.
It is important that investigators perform such simulations to determine the
method's power for their dataset, and decide if \msb has sufficient temporal
resolution to address their hypotheses; in the case of the Philippines dataset,
it did not.
When doing so, it is important to consider what prior conditions are relevant
to the empirical system.
It is rare for there to be enough \emph{a priori} information to be certain
that all taxa diverged within the last 4$\globalpopsize$ generations (i.e.,
0--1 coalescent units).
Also, it seems unlikely that when such prior information is available that
being able to detect more than one divergence event in the face of 18 random
divergences over hundreds of thousands of generations will provide much insight
into the evolutionary history of the taxa.

>>>>

\begin{linenomath}
Whereas empirical Bayesian approaches can provide powerful methods for
parameter estimation, a theoretical justification for empirical Bayesian
approaches to model choice is questionable.
In Bayesian model choice, the primary goal is not to estimate parameters, but
to estimate the probabilities of candidate models.
In a simple example with two candidate models, $M_1$ and $M_2$, we can use
Bayes' rule to calculate the posterior probability of $M_1$ as
\begin{equation}
    p(M_1 \given \alignment{i}{}) = \frac{p(\alignment{i}{} \given
    M_1)p(M_1)}{p(\alignment{i}{} \given M_1)p(M_1) + p(\alignment{i}{} \given
    M_2)p(M_2)}.
    \label{eq:bayesmodelchoice}
\end{equation}
By comparing Equations \ref{eq:bayesrule} and \ref{eq:bayesmodelchoice}, we
see fundamental differences between Bayesian parameter estimation and
model choice.
\end{linenomath}

@@

The strong preference for the narrowest prior on divergence times suggests the
approach of \citet{Hickerson2013} is biased toward models with less parameter
space and, as a consequence, will estimate model-averaged posteriors dominated
by models that exclude true values of the parameters.
We explored this possibility in two ways.
First, we re-analyzed the Philippines dataset using the model-averaging
approach of \citet{Hickerson2013}, but set one of the prior models with a
uniform prior on divergence times that is unrealistically narrow and almost
certainly excludes most, if not all, of the true divergence times of the 22
taxon pairs.
If small likelihoods of large models cause the method to prefer models with
less parameter space (Hypothesis~\ref{hypML}), we expect \msb will
preferentially sample from this erroneous prior yielding a posterior that is
misleading (i.e., the model-averaged posterior will be dominated by a model
that excludes the truth).
Second, we generated simulated datasets for which the divergence times are
drawn from an exponential distribution and applied the approach of
\citet{Hickerson2013} to each of them to see how often the method excludes the
truth.

@@

\section{The importance of power analyses to guide applications of \texttt{msBayes}}
\citet{Hickerson2013} presented a power analysis of \msb under a narrow uniform
divergence-time prior of 0--1 coalescent units ago.
They found that under these prior conditions \msb can, assuming a
per-site rate of $1.92\e{-8}$ mutations per generation, detect multiple
divergence events among 18 taxa when the true divergences were random over
150,000 generations or more.
It is important that investigators perform such simulations to determine the
method's power for their dataset, and decide if \msb has sufficient temporal
resolution to address their hypotheses; in the case of the Philippines dataset,
it did not.
When doing so, it is important to consider what prior conditions are relevant
to the empirical system.
It is rare for there to be enough \emph{a priori} information to be certain
that all taxa diverged within the last 4$\globalpopsize$ generations (i.e.,
0--1 coalescent units).
Also, it seems unlikely that when such prior information is available that
being able to detect more than one divergence event in the face of 18
divergences that were random over 150,000+ generations will
provide much insight into the evolutionary history of the taxa.
